Exploring Compiler Optimization Techniques

### 2014-11-13 doc/paper.tex
Added conclusions, register allocation implementation descriptions, and acc workings outside the acc

\section{Constant folding}
\subsection{Problem}
When a programmer writes something along these lines:

@@

\section{Constant split removal}
\subsection{Problem}
After constant folding, some \verb+split+ instructions may branch on a constant
condition:

@@

\section{Block pruning}
\subsection{Problem}
When the intermediate form is generated by the front-end, it may leave blocks in
the IR without any predecessors. This is often the result of a language that
allows the programmer to write unreachable code. I.e. the following C:

@@

\subsection{Problem}
CPUs have only a limited amount of registers. The original i386 processors had
only six general purpose registers for example (\verb+eax+, \verb+ebx+, \verb+ecx+,
\verb+edx+, \verb+esi+ and \verb+edi+).\cite[section~3.4.1]{intel}

@@

\newpage
\appendix
\section{Implementation Details for acc}
\subsection{Introduction}
  \textit{acc} (the antonijn/Antonie C Compiler) is a software project with the intent of
one-day being self-hosting (able to compile itself). The only external library
it depends on is the C99 standard library, and it's written in portable standard C99.

@@

>>>>

\section{Constant folding}
\label{sec:cfld}
\subsection{Problem}
When a programmer writes something along these lines:

@@

\section{Constant split removal}
\label{sec:uncsplit}
\subsection{Problem}
After constant folding, some \verb+split+ instructions may branch on a constant
condition:

@@

\section{Block pruning}
\label{sec:prune}
\subsection{Problem}
When the intermediate form is generated by the front-end, it may leave blocks in
the IR without any predecessors. This is often the result of a language that
allows the programmer to write unreachable code. I.e. the following C:

@@

\subsection{Problem}
CPUs have only a limited amount of registers. The original i386 processors had
only six general purpose registers for example (\verb+eax+, \verb+ebx+, \verb+ecx+,
\verb+edx+, \verb+esi+ and \verb+edi+).\cite[section~3.4.1]{intel} It is
therefore required that these registers are used cleverly to house variables\footnote{The
results of SSA instructions aren't variables strictly speaking, but for the sake
of this section they will be regarded as variables.}, so
as little as possible performance is lost due to the inherent slowness of stack
memory.

\subsection{Implementation}
\subsubsection{Restricter}
First the IR is handed over to a target-specific \textit{restricter}. This
restricter makes sure the IR meets certain platform requirements (i.e.
putting an integral/pointer return value in \verb+eax+ on x86 systems).
Making sure register requirements are met are done with the \verb+mov+ instruction.
The \verb+mov+ instruction copies another instruction result, but has a separate
life, and therefore lifetime. The \verb+mov+ result is \textit{tagged}
(annotated in the IR) with a location (\verb+loc+) tag, specifying the register.

\subsubsection{Register allocator}
To make a selection of variables to which registers can be assigned, variable
usage frequency is the most important factor. It is therefore important to
have an analysis pass analyzing variable usage frequency. A simple technique
is to simply count the amount of instructions in which the variable is used. This
disregards any control flow (and therefore block execution repetition). Such
an analysis is trivial.\footnote{This is the a\_used() analysis in analyze.c.}

A variable's lifetime needs to be considered as well. After all, no two
variables that are alive simultaneously may be assigned the same register. The
lifetime analysis step is a bit more involved, but relies on the basic principle
that a variable's lifetime starts at definition, and ends at its last usage. Its
last usage is the usage after which control can't ever reach a usage
again.\footnote{This is the a\_lifetime() analysis in analyze.c.}

Consider variable lifetimes for this block:

\begin{lstlisting}
L0:	int t1 = mul((int)2, (int)3);
	int t2 = add(t1, (int)1);
	int t3 = div(t1, t2);
	int t4 = mul(t1, t2);
	int t5 = sub(t3, t4);
	ret(t5);
\end{lstlisting}

By the target (x86) restricter, the following restrictions are made:

\begin{lstlisting}
L0:	int t1 = mul((int)2, (int)3);
	int t2 = add(t1, (int)1);
	int t3 = div(t1, t2);
	int t4 = mul(t1, t2);
	int t5 = sub(t3, t4);
	/* loc(r0) */
	int t6 = mov(t5);
	ret(t6);
\end{lstlisting}

During lifetime analysis, the following observations are added:

\begin{lstlisting}
L0:	int t1 = mul((int)2, (int)3);
	int t2 = add(t1, (int)1);
	int t3 = div(t1, t2);
	/* endlife(t1, t2) */
	int t4 = mul(t1, t2);
	/* endlife(t3, t4) */
	int t5 = sub(t3, t4);
	/* loc(r0), endlife(t5) */
	int t6 = mov(t5);
	/* endlife(t6) */
	ret(t6);
\end{lstlisting}

Such lifetime observations may be graphed into the overlap graph as shown in
figure~\ref{fig:ovl}.

\begin{figure}[h]
\begin{center}
\includegraphics[resolution=128]{ovldump.png}
\caption{Variable overlap graph. An arrow indicates a lifetime overlap.}
\label{fig:ovl}
\end{center}
\end{figure}

From this, certain registers can be induced further. There is no overlap
between \verb+t6+ and the variable it \verb+mov+s (\verb+t5+). Therefore
\verb+t5+ is also hinted to be placed into \verb+r0+.

All variables hinted towards being placed in a certain location are put there
when possible. Sometimes hints may overlap, in which case the most frequently
used is picked. The situation becomes as shown in figure~\ref{fig:ovl_ind}.

\begin{figure}[h]
\begin{center}
\includegraphics[resolution=128]{ovldump_ind.png}
\caption{Variable overlap graph after induction.}
\label{fig:ovl_ind}
\end{center}
\end{figure}

The remaining registers are allocated per variable in lexical order. If there
aren't enough registers available, the least frequently used are stored in memory.
Here, \verb+t1+ and \verb+t4+ are put in \verb+r0+, \verb+t2+ is put in \verb+r3+
and \verb+t3+ is put in \verb+r1+. These registers are \verb+eax+, \verb+edx+ and
\verb+ebx+ respectively. \verb+r3+ is chosen instead of \verb+r1+, because
\verb+r1+ is callee-save in the used calling convention (GNU i386). Caller-saved
registers are preferred over callee-saved registers because they have less
function prologue/epilogue overhead: they needn't be stored when entering the
function and restored when leaving the function.

The following tags are the result of register allocation based on these principles:

\begin{lstlisting}
L0:	/* loc(r0) */
	int t1 = mul((int)2, (int)3);
	/* loc(r3) */
	int t2 = add(t1, (int)1);
	/* loc(r1) */
	int t3 = div(t1, t2);
	/* endlife(t1, t2), loc(r0) */
	int t4 = mul(t1, t2);
	/* endlife(t3, t4), loc(r0) */
	int t5 = sub(t3, t4);
	/* loc(r0), endlife(t5) */
	int t6 = mov(t5);
	/* endlife(t6) */
	ret(t6);
\end{lstlisting}

The step of generating assembly is now only trivial.

\section{Conclusion}
There are many ways IR can be optimized into a faster form. These have been
examples of ways to do so, there are, however, many more. Most of these
optimizations have seen a practical implementation in the included \textit{acc}
software, for reference and practical examples. In appendix~\ref{sec:accdet} parts
of this software are described in detail.

@@

\newpage
\appendix
\section{Implementation Details for acc}
\label{sec:accdet}
\subsection{Introduction}
  \textit{acc} (the antonijn/Antonie C Compiler) is a software project with the intent of
one-day being self-hosting (able to compile itself). The only external library
it depends on is the C99 standard library, and it's written in portable standard C99.

@@

\subsubsection{Type system}
The type system used by the IR is the same as the type system used by C. Few
details are of importance, but it's important that the primitive types are
represented by \verb+&cint+, \verb+&cshort+, \verb+&clong+, \verb+&cchar+,
\verb+&cbool+, \verb+&cfloat+ and \verb+&cdouble+. Type system types are of type
\verb+struct ctype *+.

\subsection{Expression constructors}
\subsubsection{Instructions}
Writing instructions to a basic block is done with an instruction constructor,
which creates an instruction and inserts it at the end of a basic block. The
instruction constructor prototype for \verb+add+ is as follows:

\begin{lstlisting}
struct itm_instr *itm_add(
	struct itm_block *parent,
	struct itm_expr *left,
	struct itm_expr *right);
\end{lstlisting}

\subsubsection{Literals}
Creating a literal is done by invoking \verb+new_itm_literal()+:

\begin{lstlisting}
struct itm_literal *new_itm_literal(
	struct itm_container *c,
	struct ctype *type);
\end{lstlisting}

After creating a new literal, its value is manipulated by setting the
\verb+value+ field. The \verb+c+ parameter is needed to register the literal
with a container, to automatically dispose
of the literal when the container is disposed of.

\subsubsection{Example}
To add an addition of literals \verb+1+ and \verb+2+ (\verb+int+s) to block
\verb+b+, one'd write:

\begin{lstlisting}
struct itm_literal *l, *r;
l = new_itm_literal(b->container, &cint);
r = new_itm_literal(b->container, &cint);
l->value.i = 1;
r->value.i = 2;

/*
 * &l->base is used instead of just l
 * because the constructor takes an
 * itm_expr *, not an itm_literal *.
 */
itm_add(b, &l->base, &r->base);
\end{lstlisting}

\subsection{Optimizations and analyzations}
Optimizations as described in sections~\ref{sec:ssa}, \ref{sec:cfld},
\ref{sec:uncsplit} and \ref{sec:prune}. Are implemented in \textit{itm/opt.c}.
They are the \verb+o_phiable()+, \verb+o_cfld()+, \verb+o_uncsplit()+ and
\verb+o_prune()+ functions respectively. SSA-capability, lifetime and usage
analyses are implemented in \textit{itm/analyze.c} as \verb+a_phiable()+,
\verb+a_lifetime()+ and \verb+a_used()+ respectively.

\subsection{Command line options}
Invoke \verb+acc --help+ to obtain information about command line options. Most
importantly, to dump the IR of the C file \verb+test.c+ into \verb+test.c.ir+, run:

\begin{lstlisting}
$ ./acc test.c -Sir
\end{lstlisting}

Add \verb+-O2+ to optimize.

Using \verb+-S+ is very unstable as of the pws-bo2 version.

\subsection{Conclusion}
This has been an insight into the inner workings of acc. The optimizations
and analyzations have been left undescribed, but have been (hopefully) written
in such a way that they can be understood easily given the information provided
before.
